Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-12-28 paper2.tex
initial draft of chapters
acknowledgements

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Emulation of tree-level $e^{+}e^{-}$ matrix elements}
\label{chapter:fame1}

\end{document}

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Emulation of $e^{+}e^{-}$ to hadrons NLO k-factors}
\label{chapter:fame2}

\section{Motivation}
In the previous chapter we described in detail the strategy for emulating
tree-level $e^{+}e^{-}$ annihilation matrix elements by exploiting the factorisation
property of matrix elements. The inclusion of the dipole functions in the ansatz
meant that the infrared divergent behaviour did not have
to be learnt solely by the neural network itself, instead the task was reduced
to learning the more well-behaved coefficients of the dipole functions.

A natural extension to the work presented in the previous chapter is
to examine the emulation of the one-loop matrix elements for the same family of processes.
In general, loop matrix elements are computationally much more expensive than
their tree-level counterparts, meaning that the potential time saved by using
a fast and accurate emulator over a more traditional one-loop provider is much higher
than in the tree-level case.
In this chapter we explore this avenue by using a similar procedure to Chapter~\ref{chapter:fame1}
to construct a factorisation-aware model in order to emulate NLO QCD k-factors.

The layout for this chapter is as follows. In Section~\ref{sec:1L_fitting} we detail
the fitting procedure. Namely, we briefly recap antenna functions, as well as describe
factorisation of matrix elements within the antenna function framework in order to
arrive at an ansatz of the NLO k-factor. The construction of the neural network emulator
is described in Section~\ref{sec:1L_nn}, along with the process of training, testing and
deploying the model. In Section~\ref{sec:1L_results} we present the results of the
emulator demonstrating its speed and accuracy, where we see significant speed gains
whilst keeping the accuracy to the 1\% level. We conclude
this chapter by summarising in Section~\ref{sec:1L_conclusion}.

\section{Fitting framework for one-loop matrix element}\label{sec:1L_fitting}

In this chapter we consider the emulation of one-loop 
$e^+e^-\rightarrow Z / \gamma^{*} \rightarrow q\bar q +n_{g} g$
matrix elements for $n_{g}$ up to and including 3, which corresponds
to events with up to 5 partons. We denote the number of final-state
partons as $n$. Instead of emulating the matrix element itself we build
a surrogate for the related so-called $k$-factor
%More specifically, we consider the emulation
%of the one-loop k-factor
\begin{equation}
    k_{n} = \dfrac{2 \Re \left\{ \mathcal{M}^{(n, 0)} \mathcal{M}^{(n, 1)\, *} \right\} }{ |\mathcal{M}^{(n, 0)}|^{2} } \equiv \dfrac{|\mathcal{M}^{(n, 1)}|^{2}}{|\mathcal{M}^{(n, 0)}|^{2}} \, ,
    \label{eq:k_factor}
\end{equation}
where $\mathcal{M}^{(n, \ell)}$ denotes the amplitude
for a process with $n$ final-state partons, at loop-order
$\ell$. We will refer to the interference term in the numerator
as the one-loop matrix element henceforth and introduce
this notation for brevity.
The sum/averaging over colour and helicity is implicit for
both the one-loop matrix element and the tree-level
matrix element. The numerator in Eq. (\ref{eq:k_factor}) is the
finite part of the interference between the tree-level and 
one-loop level matrix element, where the conventional dimensional
regularisation (CDR) scheme is used.

We choose to emulate the k-factor instead of the one-loop matrix
element directly because the division of the tree-level matrix
element cancels the infrared divergences occurring for soft
and collinear external particles. However, there are still
logarithmic divergences that remain from the loop integral. Another 
advantage is that the scale of the k-factors is naturally of
the order unity, making it more amenable for emulation.

In the following, we describe how we apply the same approach as the
factorisation-aware formalism introduced in Chapter~\ref{chapter:fame1}
to encapsulate the more complex structure of the one-loop
matrix element to construct an accurate emulator for the
one-loop k-factors that is robust against single collinear or soft divergences.

An additional complication with one-loop matrix elements
is that they are evaluated at a given renormalisation scale.
This dependence can be derived from first principle, but we choose
to instead incorporate this dependence into our neural network emulator
as an input.
This method, so-called parametric
neural networks, has been utilised in other contexts
\cite{Baldi:2016fzo,Ghosh:2021roe}.

\subsection{Antenna functions}\label{sec:1L_antenna}
In building an emulator for tree-level matrix elements in Chapter~\ref{chapter:fame1},
Catani-Seymour dipoles \cite{Catani:1996vz} are sufficient
to explain all single divergences arising in phase-space. For
one-loop matrix elements we utilise antenna functions (see Section~\ref{sec:antenna_functions})
which fulfil a similar purpose of providing a set of functional behaviours
to build the matrix element out of.

Antenna functions as given in Ref.~\cite{Gehrmann-DeRidder:2005btv}
are derived from physical colour-ordered matrix elements and
by construction have the correct infrared behaviour when specific sets of
particles become unresolved. For our purposes we require the set of 
antenna functions describing the scenario of one particle
becoming unresolved at tree-level and one-loop level. Namely,
following the notation of Ref.~\cite{Gehrmann-DeRidder:2005btv},
these are the leading colour three-parton antenna functions
$X_{ijk}^{0}$ and $X_{ijk}^{1}$, respectively. $X_{ijk}^{0}$
describes all configurations where parton $j$ becomes unresolved,
where $i$ and $k$ are the hard partons. The one-loop counterpart,
$X_{ijk}^{1}$, correctly reproduces the single soft and collinear
singularity structure appearing in the one-loop singular functions
(see Section~\ref{sec:OL_factorisation}).
From this description it is clear that the antenna functions
depend on the momenta in the full $(n+1)$-body phase-space.

In one antenna function, there are two hard partons
which can both radiate off one unresolved parton. This is in
contrast to the dipole function which only has an unresolved
parton emitting from one parton. In that sense, a single antenna
function is a linear combination of two dipole functions where
the emitter and spectator are swapped. The advantage of this is
that there are generally fewer antenna functions to consider,
especially when the multiplicity increases.

Although the one-loop matrix elements data that we use for fitting are not
colour-ordered, the antenna functions nevertheless provide
a set of useful functions that allow the neural network emulator
to form accurate approximations of the one-loop k-factor.

Since we are emulating the finite part of the one-loop matrix
element we need to take care to extract all the finite parts
from the one-loop antenna functions $X_{ijk}^{1}$. The
full expression for the one-loop antenna we use is given as
\begin{equation}
    X_{ijk}^{1, F}  = \mathcal{F}inite(X_{ijk}^{1}) + \dfrac{11}{6}\log\left(\dfrac{\mu_{R}^{2}}{s_{ijk}} \right) X_{ijk}^{0} + \mathcal{F}(\mathbf{I}_{ij}^{(1)}(\epsilon, s_{ij})) X_{ijk}^{0} \, .
    \label{eq:finite_X31}
\end{equation}
where the superscript $F$ in $X_{ijk}^{1,F}$ denotes the
one-loop antenna function with all finite parts extracted\footnote{Henceforth we refer to $X_{ijk}^{1,F}$ as the
one-loop antenna function, unless explicitly stated otherwise.}.
Most of the finite parts of the antenna function are extracted in the term
$\mathcal{F}inite(X_{ijk}^{1})$. The second term
adjusts the renormalisation scale of the antenna function from
the invariant mass of the antenna partons, $s_{ijk} = s_{ij} + s_{ik} + s_{jk}$,
to the renormalisation scale the one-loop matrix element is
evaluated at, $\mu_{R}^{2}$. The final term extracts the
remaining finite parts from the singularity operators,
where their expressions are explicitly given below for
different partonic splittings.
The finite part of the singularity operators can be written as
\begin{equation}
    \mathcal{F}(\mathbf{I}^{(1)}_{ij}(\epsilon, s_{ij})) = \epsilon_{0} + \epsilon_{1} \Re{(z)} + \dfrac{1}{2}\epsilon_{2} \Re{(z^{2})} \, ,
\end{equation}
where
\begin{equation}
    z = \mathrm{zlog}(\mu_{R}^{2}) - \mathrm{zlog}(-s_{ij})
\end{equation}
and coefficients are given as
\begin{align}
    \epsilon_{0} &= \dfrac{\pi^{2}}{24} \, , \nonumber \\
    \epsilon_{1} &=
    \begin{cases}
        -\frac{5}{6}, & \text{if } ij = qg \; \text{or} \; gq \\
        -\frac{3}{4}, & \text{if } ij = qq \\
        -\frac{11}{12}, & \text{if } ij = gg 
    \end{cases} \\
    \epsilon_{2} &= -\dfrac{1}{2} \, . \nonumber
\end{align}
The $\mathrm{zlog}$ function extends the logarithm for all real-values
\begin{align}
    \mathrm{zlog}(x) = 
    \begin{cases}
        \log(x), & \text{if } x \geq 0 \\
        \log(|x|) - i \pi , & \text{otherwise} \, .
    \end{cases}
\end{align}

\subsection{Factorisation of matrix elements}\label{sec:1L_me_factorisation}
In the following, we recap the factorisation of matrix
elements as seen in Sections~\ref{sec:me_factorisation} and \ref{sec:OL_factorisation}
within the antenna function framework.

Tree-level matrix elements in $(n+1)$-body phase-space
can be factorised in the single soft and collinear limits as
\begin{equation}
    |\mathcal{M}^{(n+1, 0)}|^{2} \longrightarrow X_{ijk}^{0} |\mathcal{M}^{(n, 0)}|^{2} \, ,
    \label{eq:tree_factorisation}
\end{equation}
where $|\mathcal{M}^{(n, 0)}|^{2}$ is the reduced matrix element
in $n$-body phase-space and $X_{ijk}^{0}$ is the three-parton tree-level
antenna function
introduced in Section~\ref{sec:antenna_functions}.
The one-loop matrix element similarly
exhibits factorisation in the soft and collinear limits.
This has been extensively studied \cite{Bern:1994zx,Bern:1995ix,Bern:1999ry,Kosower:1999xi,Kosower:1997zr}
with the splitting kernels computed \cite{Kosower:1997zr,Kosower:2002su,Gehrmann-DeRidder:2005btv}.
Schematically, in the single soft and collinear limits, the one-loop
matrix element can be deconstructed into
\begin{equation}
    |\mathcal{M}^{(n+1, 1)}|^{2} \longrightarrow X_{ijk}^{0} |\mathcal{M}^{(n, 1)}|^{2} + X_{ijk}^{1,F} |\mathcal{M}^{(n, 0)}|^{2} \, ,
    \label{eq:ol_factorisation}
\end{equation}
where $X_{ijk}^{1,F}$ is the three-parton one-loop antenna function.
This equation can be thought of as a tree-level splitting
kernel multiplied by a one-loop reduced matrix element,
plus a one-loop splitting kernel multiplied by a tree-level
reduced matrix element. This is illustrated
pictorially in Figure~\ref{fig:1L_ol_factorisation}.
\begin{figure}
    \includegraphics[width=\linewidth]{fame2/oneloop_factorisation.pdf}
    \caption{Diagram illustrating factorisation of one-loop
    matrix element.
    In a singly unresolved limit, the $(n+1)$-body one-loop matrix
    element tends to the sum of a $n$-body one-loop matrix element
    multiplied by a tree-level splitting kernel and an $n$-body tree-level
    matrix element multiplied by a one-loop splitting kernel.
    The tree-level elements are drawn as a fully filled in
    circle, while the one-loop elements are drawn as a donut shape.}
    \label{fig:1L_ol_factorisation}
\end{figure}

\subsection{Ansatz for the k-factor}\label{sec:1L_ansatz}
Given that both the tree-level and one-loop level matrix
element factorise in the soft and collinear limits,
we can rewrite the $(n+1)$-body k-factor in these limits as
\begin{align}
    k_{n+1} &\longrightarrow \dfrac{X_{ijk}^{0} |\mathcal{M}^{(n, 1)}|^{2} + X_{ijk}^{1,F} |\mathcal{M}^{(n, 0)}|^{2}}{X_{ijk}^{0} |\mathcal{M}^{(n, 0)}|^{2}} \nonumber \\
    k_{n+1} &\longrightarrow \dfrac{|\mathcal{M}^{(n, 1)}|^{2}}{|\mathcal{M}^{(n, 0)}|^{2}} + \dfrac{X_{ijk}^{1,F}}{X_{ijk}^{0}} \\
    k_{n+1} &\longrightarrow k_{n} + \dfrac{X_{ijk}^{1,F}}{X_{ijk}^{0}} \nonumber \, ,
    \label{eq:k_factorisation}
\end{align}
where we see that the k-factor tends to a sum
of the reduced k-factor, $k_{n}$, and a ratio
of antenna functions. By summing over ratios of
antenna functions for all limits of a
given process, we can construct an ansatz for the
k-factor over all of phase-space.

This informs our ansatz for the k-factor, which can is given as
\begin{equation}
    k_{n+1} = C_{0} + \sum_{\{ijk\}} C_{ijk}\dfrac{X_{ijk}^{1,F}}{X_{ijk}^{0}} \,
    \label{eq:1L_ansatz}
\end{equation}
where $C_{0}$ and $C_{ijk}$ are coefficients fitted by the neural network.
$C_{0}$ is an additive term aiming to model the reduced
k-factor, and $C_{ijk}$ are multiplying the ratio of
antenna functions to fit the single collinear, and
soft limits in multiple regions of phase-space.
The sum over $\{ijk\}$ denotes the sum over
the relevant permutations of final-state partons, accounting
for all the relevant singular limits of the k-factor.
This sum allows the neural network to make use of all
the provided antenna functions to make an approximation
of the colour-summed matrix element.
The full set of antenna functions which we implement into our
model is detailed below.

The antenna functions are generically written as $X_{ijk}^{\ell}$,
which in practice is replaced with specific antennae containing either
$qg\bar{q}$, $qgg$ ($\bar{q}gg$), or $ggg$, which are
referred to as $A$, $D$, and $F$ antennae, respectively. The antennae
listed in Table~\ref{table:1L_antenna_functions} are sufficient to
describe all infrared singularities in the partonic
processes we consider.
\begin{table}
    \centering
    \begin{tabular}{cccc}
        \toprule
        $n$ & Tree-level antenna & One-loop antenna & {\{i,j,k\}} permutations \\
        \midrule
        3 $(q \bar{q} g)$     & $A_{3}^{0}(q, g, \bar{q})$ & $A_{3}^{1}(q, g, \bar{q})$  & (1, 3, 2) \\
        \midrule
        4 $(q \bar{q} g g)$   & \makecell{$A_{3}^{0}(q, g, \bar{q})$ \\ $D_{3}^{0}(q, g, g)$} & \makecell{$A_{3}^{1}(q, g, \bar{q})$ \\ $D_{3}^{1}(q, g, g)$} & \makecell{(1, 3, 2), (1, 4, 2) \\ (1, 3, 4), (2, 3, 4)} \\
        \midrule
        5 $(q \bar{q} g g g)$ & \makecell{$A_{3}^{0}(q, g, \bar{q})$ \\ $D_{3}^{0}(q, g, g)$ \\ $F_{3}^{0}(g, g, g)$} & \makecell{$A_{3}^{1}(q, g, \bar{q})$ \\ $D_{3}^{1}(q, g, g)$ \\ $F_{3}^{1}(g, g, g)$} & \makecell{(1, 3, 2), (1, 4, 2), (1, 5, 2) \\ (1, 3, 4), (1, 3, 5), (1, 4, 5), \\ (2, 3, 4), (2, 3, 5), (2, 4, 5), \\ (3, 4, 5)} \\
        \bottomrule
    \end{tabular}
    \caption{List of antenna functions we use for each process,
    and the full list of {\{i,j,k\}} permutations, where $q=1$,
    $\bar{q}=2$, $g=3,4,5$.}
    \label{table:1L_antenna_functions}
\end{table}

Since the k-factor has infrared divergences arising from
unresolved partons in the final-state being removed, and with
the appropriate antenna functions accounting for the
logarithmic divergences from the loop momenta, the
challenging task of fitting a rapidly varying function
over phase-space is reduced to fitting a group of
well-behaved coefficients that dictate how to suitably utilise
the antenna functions.

\section{Building the neural network emulator}\label{sec:1L_nn}
\subsection*{Dataset generation}
Phase-space is sampled uniformly using the {\RAMBO} algorithm
\cite{Kleiss:1985gy} with a centre-of-mass energy
$\sqrt{s_{\mathrm{com}}} = 1000$ GeV. The global phase-space
generation cut is set to $y_{\mathrm{cut}} = 0.0001$.
%We have seen that the factorisation-aware method is
%relatively agnostic to this choice of cut \cite{Maitre:2021uaa}.
We have shown in \cite{Maitre:2021uaa} that the accuracy of the emulation
is not greatly affected by the generation cut, but the extrapolation
performance is increased with a more inclusive cut, so we have chosen this
value. 
{\FastJet} \cite{Cacciari:2011ma,noel_dawe_2021_4446849} is
used to exclusively ($d_{\mathrm{cut}} = 0.01 \times s_{\mathrm{com}}$)
cluster final-state jets with the $e^{+}e^{-}$ $k_{t}$
algorithm such that there is at most a single unresolved parton.

For each phase-space point generated, we sample a
renormalisation scale, $\log(\mu_{R})$, from a uniform distribution
with end points at $[\log(s_{\mathrm{com}}/4), \log(4s_{\mathrm{com}})]$.
In other words, we sample the renormalisation scale logarithmically.
We observe that the neural network manages to learn the
renormalisation scale dependence well therefore opt to
sample $\mu_{R}$ in a wider range than is usually used
for the conventional scale variations which varies $\mu_{R}$
up and down by factors 2.

Generated phase-space points are fed to {\MadGraph} \cite{Alwall:2014hca,Hirschi:2011pa}
to compute the tree-level and one-loop level matrix elements.
For each phase-space point, we use the corresponding
renormalisation scale sampled and evaluate the strong coupling
constant at this scale using the NNPDF-4.0 NNLO PDF set
\cite{NNPDF:2021njg} with the LHAPDF6 interface \cite{Buckley:2014ana}.
All external particles are treated as massless and $m_{Z} = 91.188$ GeV.

We generate 1100k data points in total, using 100k points for
training and validation, leaving 1 million points
for independent evaluation of model performance.\
Training on a limited dataset is a realistic scenario
for processes which are prohibitively expensive, and we show
that it is possible to build an accurate emulator with the relatively
small number of data points. Note that because we are
sampling $\mu_{R}$ along with phase-space simultaneously,
there is an extra dimension in the sampled space. This means that the 100k
training points we have are not comparable to 100k training points
if we had not sampled over $\mu_{R}$. In practice, we have found that
sampling over $\mu_{R}$ has a small impact on accuracy but
opt to go this route to have the flexibility to predict over
a range of $\mu_{R}$.

\subsection*{Inputs to model}
As inputs to the neural network we provide the 4-momenta of all
final-state partons. The renormalisation scale enters the network
as $\log(\mu_{R})$ as we expect the dependence on $\mu_{R}$
to be in the form of a logarithm. Following Ref.~\cite{Maitre:2021uaa}
we include the phase-space mapping variables to aid the network in learning
the reduced matrix element information. Namely, we include as
inputs $r$ and $\rho$ from Ref.~\cite{Kosower:1997zr}
\begin{align}
    r_{ijk} &= \dfrac{s_{jk}}{s_{ij} + s_{jk}} \, , \\
    \rho_{ijk} &= \sqrt{1 + \dfrac{4r(1-r)s_{ij}s_{jk}}{s_{ijk}s_{ik}}} \, ,
    \label{eq:map_variables}
\end{align}
where $i$ and $k$ are the hard radiating partons, and $j$ is
the unresolved parton. We include the subscript
$_{ijk}$ on $r$ and $\rho$ to represent the explicit
dependence on the specific set of momenta used to calculate them.
To improve training, we transform these variables as
$r \rightarrow \log(r)$ and $\rho \rightarrow \log(\rho - (1 - \varepsilon))$,
where $\varepsilon$ is a small constant added to improve the
numerical stability. It is taken to be $\varepsilon = 10^{-8}$.
An additional input that we have observed to increase
accuracy of the emulator are the Mandelstam invariants.
These are fed into the model as $\log(s_{ij})$ for all
pairs of final-state particles. All inputs are
standardised to zero mean and unit variance.

\subsection*{Outputs of model}
The outputs of the neural network are the fitted coefficients $C_{0}$ and $C_{ijk}$
in Eq. (\ref{eq:1L_ansatz}), which when combined with the antenna functions
produces an approximation of the k-factor. We then recover the
one-loop matrix element by multiplying by the corresponding
tree-level matrix element. We do not provide the tree-level matrix
element in the emulator as the evaluation time is generally
much lower than that of the one-loop matrix element, and it is usual 
for one-loop matrix elements to be evaluated at phase-space point sets
where the tree-level matrix element has already been unweighted, so that 
only the k-factor is required.

To train the network we compare the target k-factors from {\MadGraph}
to the predictions from the neural network in the loss function by
providing the network with the antenna functions.
The target distribution is standardised to zero
mean and unit variance. Since the k-factors are of order unity
we do not need to do any additional pre-processing to aid the network
in training.

\subsection*{Neural network architecture}
\begin{figure}
    \includegraphics[width=\linewidth]{fame2/nn_architecture.pdf}
    \caption{A schematic diagram of the neural network emulator.
    The emulator is a dense neural network with inputs:
    phase-space points, $p$, momenta mapping variables,
    $r_{ijk}$ and $\rho_{ijk}$, kinematic invariants, $s_{ij}$,
    and renormalisation scale $\mu_{R}$.
    The outputs of the network are the fitted coefficients, $C_0$ and $C_{ijk}$,
    as given in Eq. (\ref{eq:1L_ansatz}).}
    \label{fig:1L_nn}
\end{figure}
A schematic of the neural network model is given in Figure~\ref{fig:1L_nn}.
We build the neural network emulator with {\Keras}
\cite{chollet2015keras} and {\TensorFlow} \cite{tensorflow2015-whitepaper}.
The emulator is a dense neural network with
three hidden layers of 64 nodes each. This network
size was chosen with consideration given to the number
of training samples and to reduce the discrepancy between training
and validation loss
(i.e. a larger network is more prone to overfitting on the
training set if there is insufficient data to fit the additional weights).
For the remaining hyperparameters we summarise the network
architecture in Table~\ref{table:parameters}.
\begin{table}
    \caption{Hyperparameters of the neural network and their values.}
    \begin{center}
        \begin{tabular}{ll}
            \toprule
            Parameter & Value \\
            \midrule
            Hidden layers & 3 \\
            Nodes in hidden layers & [64, 64, 64] \\
            Activation function & $\mathrm{swish}$ \cite{Hendrycks2016BridgingNA,DBLP:journals/corr/abs-1710-05941}\\
            Weight initialiser & Glorot uniform \cite{pmlr-v9-glorot10a} \\
            Loss function & MAE (k-factor), MSE (one-loop matrix element) \\
            Batch size & 256 \\
            Optimiser & Adam \cite{Kingma:2014vow} \\
            Learning rate & $10^{-3}$ \\
            Callbacks & \makecell[l]{{\EarlyStopping}, {\RatioEarlyStopping} \\ {\ReduceLROnPlateau}} \\
            \bottomrule
        \end{tabular}
    \end{center}
    \label{table:parameters}
\end{table}

We choose to use the mean absolute error (MAE) as the loss
function for training because it is precisely the error measure we
would like to minimise.
The error for one prediction is given as
\begin{equation}
    k_{\mathrm{true}} - k_{\mathrm{pred}} = \dfrac{|\mathcal{M}^{(n, 1)}|^{2}_{\mathrm{true}} - |\mathcal{M}^{(n, 1)}|^{2}_{\mathrm{pred}}}{|\mathcal{M}^{(n, 0)}|^{2}_{\mathrm{true}}} = \Delta \, ,
    \label{eq:mae}
\end{equation}
where the error in the one-loop matrix element normalised
by the tree-level matrix element is what we want the
neural network to minimise.
Since k-factors are a ratio of matrix elements, the numerical
values it can take are not unique for a given value of the
tree-level matrix element and/or one-loop matrix element.
For example, for two similar values of the k-factor, the scales
of the matrix elements going into each ratio may be vastly different.
To ensure that the network remains accurate for large values
of the tree-level matrix element, where corresponding corrections
contribute more to the total cross-section, we weight the training
points by
\begin{equation}
    w_{i} = \log\left( \dfrac{|\mathcal{M}^{(n, 0)}|^{2}_{i}}{\min(|\mathcal{M}^{(n, 0)}|^{2}_{i})} \right) \, ,
\end{equation}
where the $i$ index denotes training samples.
% as this minimises the total error in the combination
% $|\mathcal{M}^{(n,0)}|^{2} + |\mathcal{M}^{(n,1)}|^{2}$.
Although we use the MAE on the k-factors as the training loss,
we terminate model training based on the one-loop matrix
element accuracy. This is done by monitoring the mean squared
error (MSE) between the model prediction and corresponding truth value
at the end of each training epoch.
This takes advantage of the compact k-factor distributions
for training purposes, but bases model selection on the
accuracy for the physical one-loop matrix elements.

To reduce the effects of overfitting we have two {\EarlyStopping}
criteria: the first is to stop training once the validation loss has
not improved in 100 epochs, and the second is a {\RatioEarlyStopping}
which terminates training if the ratio of training loss to validation
loss drops below a certain threshold. We take this threshold
to be $0.9$. We also use the {\ReduceLROnPlateau}
callback as a way to adapt the learning rate during training.
The learning rate is reduced by a factor of 0.7 whenever
validation loss plateaus with a patience of 20 epochs. We find
that with these hyperparameters we achieve a balance of reducing
overfitting and quick training times. On average the models
train in approximately 20 minutes on an Nvidia P100 GPU.

Although we build and train our model using {\TensorFlow},
we deploy the model using the Open Neural Network Exchange ({\ONNX})
runtime \cite{onnxruntime}
with the {\CUDA} execution provider to run predictions on an
Nvidia P100 GPU. With the optimised operations in the {\ONNX}
runtime, we see that compared to {\TensorFlow} the model inference
time is reduced by an order of magnitude or more. Another advantage
is that it gives users the flexibility to move the pipeline
away from {\TensorFlow} on $\mathtt{Python}$ to a more generic interface
to the neural network model. One example would be to integrate
the {\ONNX} model into a \texttt{C++} workflow for use with current event
generators to replace the one-loop provider with a neural
network emulator.

In addition to the {\CUDA} execution provider, we will use
the {\ONNX} runtime CPU execution provider to compare with
{\MadGraph} for a comparison of single CPU core performance.
This will be the closest to a real world benchmark as event
generators typically generate events on a single core.

Although we find that the neural network models converge well,
to account for stochasticity in the training, the random
initialisation of model parameters, and to reduce
variance on predictions, we initialise 20 models for training
and use the mean of these models as our model prediction. This
ensembling will also give a measure of the uncertainty
due to the neural network optimisation, by using the standard
deviation of the 20 replica model predictions.

We plot in Figure~\ref{fig:1L_loss_curves} the losses of the 20 replica
models for the 5 jet process, where we have plotted the loss for the
k-factor (MAE) and the one-loop matrix element (MSE).
We can see that the models have all
converged to a similar point when training is terminated. The
noise in the validation loss at the beginning of training can be
attributed to the fitting of the coefficients: when the model is
learning how to pick the relevant combination of antenna functions
there can be large variations in the prediction, however, the
variations become much smaller once the model learns the
factorisation properties and converges. We can
see the variations in the validation loss are small
at the end of training, and that there is not a large discrepancy
between the training and validation losses, as enforced by the
{\RatioEarlyStopping} callback.

\begin{figure}
    \centering
    \includegraphics[scale=0.57]{fame2/loss_curves.pdf}
    \caption{Training and validation losses of the 20 independent
    initialisations of the ensemble replicas plotted as individual
    curves for the 5 parton model. We plot the MAE loss for the k-factor
    and the MSE loss for the one-loop matrix element. The scale difference
    between the MAE and MSE losses is a consequence of the form of the loss
    functions and is not surprising.
    We see that the replica models all converge
    to a similar point with the validation loss being close to the training loss
    when training is terminated. The step feature in the training loss
    is due to the {\ReduceLROnPlateau} callback.}
    \label{fig:1L_loss_curves}
\end{figure}

\section{Results}\label{sec:1L_results}
\subsection{Error distributions}
In this section we present results for our NLO QCD k-factor
emulator for $e^{+}e^{-}$ annihilation into up to 5-jets.
First we show a comparison between our model described in
Section~\ref{sec:1L_nn} which we label `antenna',
and a `naive' model with no factorisation properties built
into the emulator: a densely connected neural network
with the parameters given in Table~\ref{table:parameters}
that directly predicts the k-factor.
For the `naive' models we train with the MAE loss on the k-factors
with no modifications and terminate training
based on this loss. As with the `antenna' models, we also
ensemble 20 individual replica models for the `naive' model predictions.
In Figure~\ref{fig:1L_delta_distribution} we compare histograms
of $\Delta$ for all final-state multiplicities between these two models.
It is immediately clear that building
in the factorisation structure of the matrix elements greatly increases
accuracy, with increasing relative improvements
for the higher multiplicity cases.
The `antenna' error distributions are symmetric, strongly peaked around
the ideal value of 0, and with tails falling off rapidly. We
see the general trend of increasing multiplicity decreases accuracy,
however we observe that the bulk of the 5 jet final-state is
within percent accuracy and with the lower multiplicities well
below this.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fame2/all_jets_ratio_comparison_plots.pdf}
    \caption{Error distribution in terms of $\Delta$ for all multiplicities. `Antenna'
    model is as described in this article, and `naive' model is a simple densely connected
    neural network model without any factorisation properties built in.
    We keep the horizontal axis scale fixed for all subplots to make it easier to compare
    accuracy across the different multiplicities.}
    \label{fig:1L_delta_distribution}
\end{figure}

To show that we are accurate across the entire span of the tree-level
matrix elements, and to have a closer inspection of the tails
of the $\Delta$ distribution, we plot a 2d histogram of $\Delta$
against the value of $|\mathcal{M}^{(5, 0)}|^{2}$ in
Figure~\ref{fig:1L_delta_hexplot}. The bulk of phase-space points
are contained in the high population bins depicted in yellow,
representing the peaked distribution seen in Figure~\ref{fig:1L_delta_distribution},
whereas the green to purple coloured bins represent the tails of the
low population $\Delta$ distribution. We see that the accuracy
stays contained inside a band and does not flare out
as the magnitude of the tree-level matrix element increases.
This shows that we manage to fit the k-factor even in
the infrared and collinear limits where the tree-level
matrix element becomes large. On the right-hand side subplot we plot
the distribution of the tree-level matrix elements where
we can see that that even with relatively few training
points in the tails, the emulator is still able to predict
these regions as well as where there is more abundant data.
\begin{figure}
    \centering
    \includegraphics[scale=0.65]{fame2/5j_hexplot.pdf}
    \caption{Left: $\Delta$ error distribution plotted against the
    tree-level matrix element. Yellow bins indicates high density regions
    of points and purple bins indicates single points.
    Right: marginal distribution of tree-level matrix elements.
    This illustrates that the network is able to reproduce
    a good approximation across all sampled phase-space.}
    \label{fig:1L_delta_hexplot}
\end{figure}

\subsection{Renormalisation scale dependence}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fame2/renormalisation_scale_trajectories_logscale.pdf}
    \caption{Renormalisation scale trajectories for all multiplicities.
    These trajectories are predictions of the k-factor with the sum
    of antenna functions subtracted to show renormalisation scale
    dependence in the remainder. Each trajectory is at one phase-space point
    sampled from {\RAMBO}, with $\mu_{R}$ spanning the range
    $[s_{\mathrm{com}}/8, 8 s_{\mathrm{com}}]$. Each trajectory
    is composed of 1000 points.
    The region that $\mu_{R}$ was uniformly sampled from
    for training is indicated as a red band. The error bands on the NN
    predictions are too small to be seen.}
    \label{fig:1L_mu_trajectory}
\end{figure}

In Figure~\ref{fig:1L_mu_trajectory} we show that our
emulator has learned the renormalisation scale dependence,
independent of the antenna functions. To produce
a trajectory we first sample a phase-space point with
the same cuts as described in Section~\ref{sec:1L_nn}, then we
evaluate the k-factor at this phase-space point with $\mu_{R}$
varying from $s_{\mathrm{com}}/8$ to $8s_{\mathrm{com}}$.
We choose to sample from a wider range than used for training
to examine the $\mu_{R}$ extrapolation performance.
After subtracting the sum of the antenna functions we see that the
remainder still has a dependence on the renormalisation
scale that is accurately captured by the neural network.
As with the $\Delta$ distribution plots in Figure~\ref{fig:1L_delta_distribution},
there is a slight decrease in accuracy as we increase the
multiplicity, however, the ratios and differences are well-behaved
throughout the entirety of the trajectories inside the
range of training data. The only anomaly occurs when the
trajectories cross zero, causing spikes in the ratios,
but the difference in truth and prediction remains well-behaved
around these regions. Outside of the training
range we see an acceptable extrapolation, but given that
the training range is wider than the range in which the
renormalisation scale is normally varied for scale
variations, we do not find this problematic.

\subsection{Total cross-section predictions}
In Figure~\ref{fig:1L_cross_section} we show that we reproduce
the total cross-sections over an integration of 1M phase-space
points, for all multiplicities, at three
different values of the renormalisation scale representing the
nominal value ($\mu_{R} = \sqrt{s_{\mathrm{com}}} = 1000$ GeV)
and the two variations usually used to estimate scale uncertainties.
Note that these phase-space points were
generated independently of those in Figure~\ref{fig:1L_delta_distribution},
and that we are integrating the tree-level and one-loop interference, not the k-factor.
We multiply our NLO k-factor prediction with the {\MadGraph} tree-level
matrix element to reproduce the loop-matrix element for integration.
The neural network errors are well
below the statistical Monte Carlo integration error.

\begin{figure}
    \includegraphics[width=\linewidth]{fame2/cs_plots.pdf}
    \caption{Error on the total cross-section across an integration
    of 1 million phase-space points. For each multiplicity, we
    evaluate the matrix elements at the three values of the 
    renormalisation scale as reported.
    The solid lines are the absolute percentage error in the true total
    cross-section and the neural network predicted value. The dashed
    line represents the statistical Monte Carlo integration error
    which falls as $1/\sqrt{N}$. The jumps in error are due to large
    values of the matrix element being integrated.}
    \label{fig:1L_cross_section}
\end{figure}

By neural network error we are referring to the absolute percentage
difference to the cross-section, and not the errors due to neural
network optimisation. For that, we examine the variations in the 20
replica model predictions in Figure~\ref{fig:1L_5j_cs} where we plot the
total cross-section predictions as a scatter plot. The blue band
illustrated is one standard deviation of the 20 predictions made.
We see that the true value of the total cross-section is within
the one standard deviation band. Not shown in the figure is the
Monte Carlo statistical error which as seen in Figure~\ref{fig:1L_cross_section}
dominates the absolute error between the NN predictions and
the true value.

\begin{figure}
    \includegraphics[width=\linewidth]{fame2/5j_cross_sections.pdf}
    \caption{Scatter plot showing the variation in predictions of
    the 5 jet total cross-section for the 20 replica models, where
    the plotted cross-section is normalised by the truth value
    provided by {\MadGraph}.
    Total cross-sections in each subplot are evaluated at the
    renormalisation scale quoted in the upper right.
    The mean prediction of the 20 replica models is drawn as the horizontal
    blue line with one standard deviation illustrated by the blue band.
    The total cross-section as predicted by {\MadGraph} is plotted
    as the horizontal black line.}
    \label{fig:1L_5j_cs}
\end{figure}

Since the discrepancy between the true total
cross-section and NN predicted total cross-section is so small,
this strongly indicates that once we fit on the relatively small
training set, we can predict with good confidence on many more
phase-space points to get a prediction of the cross-section before
reaching the same level of error as the Monte Carlo integration error.
This can also be seen from the shape of the NN error, it is relatively
constant once enough points have been integrated.
